---
title: "STAT4520 HW4"
author: Anton Yang
output: pdf_document
date: "2024-10-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1

Suppose data is generated from the exponential distribution with density
\[f(y) = \lambda e^{-\lambda y}\].
We can write this as exponential family:
\[f(y|\theta, \phi) = e^{-\lambda y + \log(\lambda)}\]

So, $\theta = -\lambda$, $\phi = 1$, $a(\phi) = \phi = 1$, $b(\theta) = -\log(\lambda)$, and $c(y, \phi) = 0$. 

We can make the exponential distribution into Canonical Form, which we let $\eta = -\lambda$. 

Therefore, $\eta = -\lambda$, $\phi = 1$, $a(\phi) = \phi = 1$, $b(\eta) = -\log(-\eta)$, and $c(y, \phi) = 0$.

Thus, we can solve $E[X]$ and $Var[X]$ with $b(\theta)$.

\[E[X] = b'(\theta) = \frac{-1}{\eta} = \frac{1}{\lambda}\]
\[Var[X] = b''(\theta) = \frac{1}{\eta^2} = \frac{1}{\lambda^2}\]

## Problem 2
```{r}
library(faraway)
library(dplyr)
library(mgcv)
data<-chicago

tw_model<-gam(involact ~ age + theft + log(income), family = tw(link = "log"), data = data)
summary(tw_model)
```

According to the summary, we have a p of 1.152 and $\phi = 0.44757$.

```{r}
xgrid<-seq(1e-10, 1.25, len = 100)
p <- 1.152
phi <- 0.44757
mu <- tw_model$fit[1]
poismean<-mu^(2-p)/(phi * (2-p))
p0<-exp(-poismean)

twden<-exp(ldTweedie(xgrid, mu, p = p, phi = phi)[,1])
data$involact[nrow(data)]
plot(xgrid, twden*(1-p0), type = "l", xlab = "x", ylab = "Density", main = "Observation 60645 Predictions")
dmax<-max(twden * (1-p0))
segments(0, 0, 0, dmax, lwd =  2)
text(0.05, dmax, paste0("p =", signif(p0, 3)))

tw_model2<-gam(involact ~ age + theft + log(income) + fire + volact, family = tw(link = "log"), data = data)
summary(tw_model2)

p2 <- 1.132
phi2 <- 0.3841
mu2 <- tw_model2$fit[1]
poismean2<-mu^(2-p2)/(phi * (2-p2))
p0_2<-exp(-poismean2)

twden2<-exp(ldTweedie(xgrid, mu2, p = p2, phi = phi2)[,1])
plot(xgrid, twden2*(1-p0_2), type = "l", xlab = "x", ylab = "Density", main = "Observation 60645 Predictions Full Model")
dmax2<-max(twden2 * (1-p0_2))
segments(0, 0, 0, dmax2, lwd =  2)
text(0.05, dmax2, paste0("p =", signif(p0_2, 3)))
```

The p value for the full model (0.416) is slightly higher than that of the smaller model (0.398). Additionally, the density plot shows that the full model has a higher maximum value, and its curve appears more irregular compared to the smoother curve of the smaller model





